### 2009-12-09 docs/marcello-dissertation/thesis/chapter2-state-of-the-art.tex
Changes to chapter 2, after Prof. Puder's and Niki's revision.
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@682 559edde2-a4d4-44e0-96fc-4035e40d7396

\chapter{State of the Art}

In general, this chapter reviews surveys and related papers in networks
networks, providing a basic introduction about their characteristics to new 
audience. Similarly, it reviews the characteristics of related works that
discusses and provides data persistence solutions for sensor networks, 
characterizing the different approaches used to described and model the
gathered data from sensor devices.

@@

According to \cite{sn-intro02}, sensor networks are specialized types of
network systems comprised of nodes with specific goals such as to produce and
collect data. The former is seen as a sensor device capable of producing data
to the network based on an observation from the environment or simply a raw
data. On the other hand, the latter is described as the final destination of
the raw data in a sensor network and called the network data sink, or simply
data sink, whose responsibility is to receive, categorize and store the
``sensed'' data. In this way, the location of the the data sink is determined by
the purpose of the collected data: data nodes are placed in-network close to
the sensor devices when collected data has a life cycle, being discarded after
being utilized. In contrast, data nodes are placed in in a centralized node
when the collected data is used for historical purposes.

@@

Each sensor node perform different tasks for the network and, therefore,
influences on the method of deployment used. \cite{sn-intro01} identifies that
the deployment of sensor devices can use arbitrary methods such as being
dropped from an airplane, or manually placed at chosen geographic locations.
Moreover, those authors also mentioned that the deployment can be done as a
one-time event or as an iterative process, where the sensor devices are moved
to different places as needed.

Moving to a differnt point, \cite{sn-intro01} reports that sensor nodes can be
able to move around the place where they were deployed or a stationary sensor
at a given location. In the case of a sensor networks comprised with different
moving sensors, a given sensor node can move occasionally or continuously, as
a result of an accidental side-effect or a desired property of the device. As
a matter of fact, these sensors are automotive or they can be attached to
another moving object. For this reason, \cite{sn-intro01} classifies those
types of mobility as active or passive.

Considering the physical design and capabilities of sensor devices, they range
from microscopically small to bigger devices of a few cubic feet
\cite{sn-intro02}. In addition to their limited size, their computational
resources and power availability are also scarse, when they are located in
areas of limited access to energy. In this way, \cite{sn-intro01} also
distinguishes how sensor devices are usually designed to either store energy on
batteries or use energy scavenged from solar cells. Even though a sensor
network is designed for a purpose, \cite{sn-intro01} also explains that it can
be comprised of different sensor devices playing different roles on what they
observe in the environment, characterizing the heterogeneity the sensor
network by the hardware capabilities and what types of data each of them
produce.

Finally, \cite{sn-intro01} mentions that sensor device nodes can communicate
with each other using different media such as radio, diffuse light or
inductive sound, influencing on the type of protocols they exchange data. They
can have knowledge of their physical location if they are integrated with a
GPS\footnote{Global Positioning System} device, which gives the latitude or
longitude coordinate values. Therefore, their cost of these devices may range
from hundreds to thousands of dollars, depends on a given device's functionality
and capabilities of its design.

@@

Last section discussed the different properties of the sensor devices when
placed in a sensor network. Depending on how they communicate with each other,
\cite{sn-intro02} discusses that sensor devices can be in an organized in a
structured or ad hoc way. For this reason, the infrastructure, of sensor
networks differ in topology such as single-hop, star, tree or  graph. Each of
these topologies may influence on how the produced data from sensor nodes is
collected and routed to their related data sink. While there are only two
levels of node connectivity in sensor networks deployed as single-hop and start
topologies, sensor networks whose topology is either tree or graph use
multi-hop communication among each of their nodes for data communication from
the devices to the data sink.

The property of coverage on a sensor network is defined by
\cite{sn-intro01} as the physical space ``visible" to the sensors devices, which
is also directly related to the purpose of the network, its size in terms of
the number of nodes deployed and the sensor network lifetime. Therefore,
sensors are deployed in sparse or dense manners, according to the
specification of what needs to be covered by the sensor devices. Furthermore,
considering that some types of sensor devices are very likely to regular
hardware failures, \cite{sn-intro02} described sensor network examples with 
redundant nodes that provide better data coverage and levels of coverage.
Finally, the coverage of sensor devices also influence on the connectivity
property of the sensor nodes. In order communicate with redundant nodes, a
physical connection must be available at all times. On the other hand, lower
level of coverage is is related sporadic communication among the sensor nodes.

The sensor network size is related to the degree of coverage mentioned earlier,
and it is measured by the number of devices deployed. \cite{sn-intro01}
reports examples of sensor networks composed by different ranges of deployed
sensor devices such as dozens, hundreds or even thousands of nodes. However,
the life time of a sensor network determines how long these devices are going
to be used, as well as how long the sensor network may exist. For example, an
environmental sensor network \cite{sn-ex01} with the purpose of collecting data
for historic purposes may be have their sensor devices continuously performing
their data collection tasks. On the other hand, however, the lifetime of a
sensor network designed to achieve a goal may not exist after the sensor
devices finish their tasks. Finally, \cite{sn-intro02} argues that the
performance related to the orchestration of sensor devices in sensor networks
is related to the constraints defined for quality-of-service (QoS) such as
real-time data delivery within a thredshold or for robustness such as
maintaining data alive even though a communication link is not available.

All things considered, sensor nodes play an important role during the execution
of a sensor network, whose infrastructure is characterized by how and where the
sensor nodes are deployed and communicate with each other. In this way, next
section covers the approaches to deal with the product of this type of system
network.

@@

As traced in the last few sections, sensor networks are comprised of
different types of sensors nodes. However, an important sensor node type is the
data sink because it is the place where the raw data produced by sensor
nodes resides. Data collection and access in sensor networks is directly
related to the characteristics of the sensor network, its routing mechanisms
and where the data sink is placed within the network. In a nutshell,
\cite{sn-storage03} describes collected data as can be described as Low-level
property or an observation such as the level of the ocean, etc.

The type of query mechanism used to directly or indirectly interrogate sensor
devices in a sensor network is directly related to the aforementioned locations
of the collected data. The in-network query processing is used when the
location of the collected data is a local storage \cite{sn-storage01,
sn-storage03} located in the network, strategy typically used to maximize the
sensor's resources, while minimizing energy consumption \cite{sn-storage04}. On
the contrary, the a typical centralized query processing is used to search for
values over the collected data when its location is a centralized data
node \cite{sn-storage02}.

When it comes to the data representation, the relational data model
\cite{relational-model} is one of the most traditional approaches used to
describe the collected data of sensor networks, as it is reported in database
systems such as the TinyDB \cite{sn-db-tinydb}. For instance, \cite{sn-db-newop}
proposes a modified version of the SQL language of TinyDB proposed to solve
SQL Join problems, so that the query processing takes into account the location
of the distributed data. Conversely, \cite{sn-provenance} proposes the use of
Data Provenance as the primary guidelines to describe collected data from
sensor devices, giving the fact that these devices only produce observations as
stream of raw data, and therefore, described to address the problems related to
the origin of the data.

@@

At first glance, the raw data serves as the main outcome of the sensor network.
\cite{sn-provenance} differentiates the use of the collected data as follows as 
\textbf{real-time data stream}, when they are generated data from a sensor
device whenever the data is requested. In this way, the data is still in its raw
format specified by the sensor manufacturer; On the other hand, the authors
describes the data as \textbf{archivable}, since they a have historical
significance and will be processed to be reused by other systems in a latter
time. While the real-time data streams are used and potentially discarded, the
archivable data is stored into secondary memory such as a hard-disk. Under
these circumstances, the collected data needs first to face its journey
starting from its production, transmission from the sensor device, to finally
reach a specified data sink in the network \cite{sn-storage01, sn-storage02,
sn-storage03}.

After producing the ``sensed'' values for the observed parameters from the
environment, the sensor node is responsible to send the raw data to a specific
data node defined by the network infrastructure, as delineated in section
\ref{sec:sn-infrastructure}. In such a way, \cite{sn-storage03}
describes that the location for the collected sensor data provides two
different types of storage devices: the \textbf{external storage} and the
\textbf{local storage}. In the former type, the sensor device sends the
produced data to another sensor node that contains an external storage, whereas
the data only exists in memory or between its reading to its delivery as
real-time data stream in the the latter type. On the whole, the sensor network
infrastructure dictates where the movement performed by the sensor data, as
\cite{sn-storage02} sees a wireless sensor networks as a many-to-one data
collection pattern, when its organization is based on single-hop or star. On
the other hand, data may also flow from one intermediate sensor node to another
until it reaches its final data sink when the data organization is structured
as a graph, tree \cite{sn-storage01, sn-storage03}.

Last, but not least, the use of different external storage devices can be used
to partition the collected data. This practice is called \textbf{Data-Centric
Storage}, and aims at clustering the collected data from different sensor
devices based on a given a property of the data such as location or time of
creation.

@@

The query processing in sensor networks is a direct or indirect method of
accessing the collected data stored at data nodes, whose location is determined
by different facts described in the previous section. \cite{sn-storage03}
describes an example of an \textbf{in-network query processing} when sensor
nodes, acting as data nodes, provides real-time data stream as a response
to queries requesting the device's collected data. Likewise, if the collected
data is in use for a specific period of time, the data is cached in an external
storage device in a close-by data node. In this way, when a query is issued to
the sensor network with these characteristics, the query processor considers 
all data nodes return its final response to the query. In comparison, a sensor
network whose topology is in form of star, the data sink is usually located in
the center of structure, and thus, receives all the collected data from other
sensor devices in the network. For this reason, it can be seen as a 
\textbf{centralized query processing}. Factoring in that a centralized data
sink receives all the collected data from the peer nodes, this query processing
strategy is the solo responsible for responding to indirect queries about the
produced data from the network.

Based on the literature review, there are advantages and disadvantages related
each of the query processing approaches described with regards to the
infrastructure of the sensor network. \cite{sn-storage03} argues that the 
primary purpose of the in-network query processing is to minimize the number
of data transmitted from sensor devices to a data sink, as sensor devices are
generally limited by energy use. Considering that sensor devices are also used
as data nodes, \cite{sn-storage04} delineate that this approach can potentially
flood all sensor nodes in the network, and therefore, decrease its performance.
For this reason, \cite{sn-storage01} reports the use of the data-centric
storage by organizing the collected data based on its particular properties
values. In this way, the query processor hits specific data nodes that contains
the types of data in the range of the requested data and, therefore, the author
concludes that this approach decreases the network congestion and maximizes the
capacity on the sensor devices and the performance of the sensor network.

Unlike the in-network, the centralized query processing is related to a regular
database system, located in one of the network nodes. For this reason, this
indirect type of query processing is used when the data purpose of the use of
data is for data archival. However, \cite{sn-storage02} showes that as a
consequence of the many-to-one communication between the centralized data sink
and the other data nodes, this query process does not affect any of the sensor
nodes in the network and, thus, maximizes the sensor device utilization. 
Moreover, this strategy better exposes the collected data to the primary users
raw data, which follows the process of data processing, compression,
aggregation, and so on \cite{sn-db-modeling02}.

Although the approach of centralized query processing maximizes the sensor
nodes capacity and minimize energy consumption, \cite{sn-storage02} explains
that this trade off imposes the creation of the unavoidable development of a
point of traffic concentration or the phenomenon of network congestion. For
this reason, \cite{sn-storage04} describes that different techniques used in 
distributed system such as the use of master-slave data replication or 
database partitioning \cite{db-partitioning-relational} can be used to reduce
or minigate the effects of this funneling effect.

@@

Different approaches are used to represent the collected data from sensors. The
use of basic data descriptors has been used in research groups which expertise
is not database systems, while others use such technology in pragmatic
approaches. \cite{sn-programming-language} reports different applications using
different types of languages to acces the data modeled in different data
models. \cite{sn-provenance} describes the use of a \textbf{Tabular Data Model}
\cite{tabular-model} using comma-separated values to model and organized the
values related to the properties of observed parameters from a given sensor
device. On the other hand, \cite{sn-db-tinydb} describes the use of the
\textbf{Relational Data Model} \cite{relational-model} is used in the majority
of the projects that involves data persistence due to its popularity in other
different areas of Software Engineering and Distributed Systems. Finally,
\cite{sn-xml-usage01, sn-xml-usage02} are examples of sensor networks
application that makes use of XML \cite{xml} to represent the collected data
on a \textbf{Structured Data Model}.

@@

The most basic computer application to process data is through the use of
spreadsheets, where data is organized into cells \cite{tabular-model}. When
exported to the so-called comma-separated values (CSV) file format, the
resulting file is comprised of lines of ascii characters, separated by a
delimiter. The first line represents a header, with the identification of
columns of specific data, where the following lines contains the values
related to each top column. \cite{sn-provenance} reports that the tabular
data model is commonly used to provide data persistence for sensor networks.
As a matter of fact, the SF-BEAMS environmental sensor networks
\cite{sfbeams2006} uses this approach to distribute the collected data over
the Internet using the OPenDAP \cite{opendap} data format.

This data model can be used in different context, as \cite{sn-provenance}
describes that the versions of the same document can be easily managed and
shared among research colleages using a version-control system such as
Subversion \cite{subversion}.

@@

\cite{sn-db-newop} shows the use of the Relational Data Model
\cite{relational-model} can be used in sensor networks to describe the
collected data from sensor devices. As a matter of fact, the project described
by the authors is based on the TinyDB database system \cite{sn-db-tinydb}, a
database system specifically built for data nodes of sensor networks. It uses
the same properties and constraints of regular database system, where the
entities of the sensor network are required to be modeled and represented by
tables.

@@

XML is an increasing data format used popularized by the applications developed
for the Internet. \cite{sn-xml-usage01} reports an efficient way to represent
the collected data from sensors using XML, while the authors of
\cite{sn-xml-usage02} describe a sensor network using XML for data model
representation. In general, the collected data is structured into classes of
relating elements and validated by an XML Schema \cite{xml-schema}.

@@

Other examples of the use of the structured data model using XML is the through
the use of middlewares, as the authors of \cite{sn-xml-query-engines} reports.
Such approach is used for data exchange by the implementing of a middleware
capable of query and extract the collected data from sensor devices
\cite{sn-xml-middleware}. A practical example of the use of XML as a middleware
to communicate between nodes is the communication mechanism used by NetBEAMS
\cite{netbeams2009}. See section \ref{sec:dsp-serialization} for details.

\section{Data Provenance: How Collected Data is Described?}
\label{sec:sn-provenance}

As described in the previous section, the collected data from sensor networks
can be represented by different data models, requiring the knowledge of the
properties of the data. Not to mention that the low-level or raw data produced
by a given sensor device might not include vital information regarding the
origing and identity of the collected data. For this reason,
\cite{sn-provenance} describes Data Provenance as a technique used to describe
those essential properties of the collected data from sensors. One typical
example of the application of Data Provenance is the so-called blackbox of an
aircraft, which carries the collection of raw data collected from a group of
sensors placed in different locations of an aircraft, as shown by
\cite{sn-exemple-blackbox}. Last, but not least, the author describes the
collected data as used in a process of data recovery and analysis in order to
identify the probable reason of an aircraft accident, by using the time, and
location of the aircraft, along with the independent variables from the
sensors.

In general, sensor devices produce data and either keep them onto its local
storage or transmit them to data nodes as described earlier in this chapter. It
is vital to understand that the majority of these devices outputs the data in
its raw format, containing only a stream of values with any type of
delimitation between each other. For this reason, the processes of parsing and
description of each of the values located in the data streams regularly require
the use of the device's specification documentation. Similarly, in order to
have access to each of the properties of data collected by a given property
related to time or space, \cite{sn-provenance} also mentions that 
Metadata\footnote{Data about data} should be included together with the
collected data.

An example of Data Provenance application is the use of aggregation of data
over time to estimate the changing effects of the weather in a given region.
For this reason, the collected data from sensor devices need meaningful
annotations to better describe the identity of the collected data, as 
\cite{sn-provenance} reports. Interesting questions regarding the
data can be asked such as \textbf{what was collected?}, \textbf{When was the
data collected?} or \textbf{Where was the data collected?}. Similarly, while
Data Provenance provides basic mechanisms to describe sensor networks data in
general, others see the importance of the time space of sensor networks.
\cite{sn-time-series} describes strategies used to cluster distributed sensor
data as time series of data, as well as to decrease the size of data streams
being transmitted in real-time sensor networks when these collected data does
not need to be transferred \cite{sn-data-reduction}.

Finally, data provenance provides practical use of the collected data from
different sensor devices and different sensor networks in the process called
Information Fusison \cite{sn-info-fusion}. An example of the aggregation of
different data from different devices located in different geographic sensor
networks is \cite{sn-geo-metadata}, which is capable of process
spatial-temporal queries of real-time or historical data. The following
sections describe each of the metadata properties as defined by
\cite{sn-provenance}, giving different examples of each of them.

\subsection{What was collected?}

Different sensor devices produce different sets of attributes based on its
capabilities and design. In general, these data sets do not have distinct
names, which leads to difficulties when indices over the properties of the
collected data are required to make them searchable. Specially for the purpose
of data archival, the collected data needs to be identified, giving a meaning
of independence on its own, the same idea behind the unique identification of
entities on the relational data model \cite{relational-model}. Equally,
\cite{sn-provenance} describes that traditional sensor networks uses the
collected data for analysis long after the attributes have been collected,
usually before it is categorized and indexed. Annotations and tags are also
names given by \cite{sn-provenance} as means to describe collected sensor data.

In a nutshell, the collected data is described by the name of the observation
property. For instance, \textbf{temperature} = 73.45 expresses the value of the
observation of the temperature. In other words, the string ``temperature'' is
the property describing the raw collected data 73.45. On the other hand,
however, the example \textbf{temperature-scale} = ``Fahrenheit": describes
the metadata regarding the temperature.

\subsection{When was the data collected?}

In order to use the collected data for the purpose of historical analysis, the
use of a point in time is the most prominent feature of data, as it is
considered a pivotal metadata for different types of sensor networks such as
an aircraft the blackbox \cite{sn-exemple-blackbox} or any
environmental sensor network \cite{sfbeams2006, sn-xml-middleware, sn-ex01}.
In this way, Other projects classify data in time-series
\cite{sn-time-series-example}, where data is related to all the
observations a given sensor device performed over a period of time. The
apparent implications is that all the produced data is necessary, but
\cite{sn-time-series} argued that some sorts of sensor networks could decrease
the creation of ``noisy data sets" based on the proper clustering of
similar time-series data before before they are sent to the data sink.

>>>>

\chapter{State of the Art of Data Persistence in Sensor Networks}

In general, this chapter reviews surveys and related papers on sensor networks,
providing a basic introduction about their characteristics to a new audience.
Similarly, it reviews the characteristics of related works that discuss and
provide data persistence solutions for sensor networks, characterizing the
different approaches used to described and model the gathered data from sensor
devices.

%It is important to note that the scope of this literature review is limited to
%our case study described in Chapter 4 and, therefore, it only
%covers a subset of disciplines in sensor networks and data persistence.

@@

According to \cite{sn-intro02}, sensor networks are specialized types of
network systems comprised of nodes with specific goals such as to produce and
collect data. The former is seen as a sensor device capable of producing data
to the network based on an observation from the environment. On the other
hand, the latter is described as the final destination of the raw data in a
sensor network and called the network data sink, or simply data sink, whose
responsibility is to receive, categorize and store the ``sensed'' data. In
this way, the location of the data sink is determined by the purpose of
the collected data: data nodes are placed in-network close to the sensor
devices when collected data has a life cycle, being discarded after being
utilized. In contrast, data nodes are placed in a centralized node when the
collected data is used for historical purposes.

@@

Each sensor node performs different tasks for the network and, therefore,
influences the method of deployment used. \cite{sn-intro01} identifies that
the deployment of sensor devices can use arbitrary methods such as being
dropped from an airplane, or manually placed at chosen geographic locations.
Moreover, the cited authors also mentioned that the deployment can be done as a
one-time event or as an iterative process, where the sensor devices are moved
to different places as needed.

As noted by \cite{sn-intro01} sensor nodes can be
moved around the deployment location, or remain in one place as a stationary sensor. 
In the case of a sensor network comprised with different
moving sensors, a given sensor node can move occasionally or continuously, as
a result of an accidental side-effect or a desired property of the device. As
a matter of fact, these sensors are automotive or they can be attached to
another moving object. For this reason, \cite{sn-intro01} classifies those
types of mobility as active or passive.

Considering the physical design and capabilities of sensor devices, they range
from microscopic to devices measuring a few cubic feet
\cite{sn-intro02}. In addition to their limited size, their computational
resources and power availability can also be affected, when they are located in
areas of limited access to energy. In this way, \cite{sn-intro01} also
distinguishes how sensor devices are usually designed to either store energy in
batteries or use energy scavenged from solar cells. Even though a sensor
network is designed for a purpose, \cite{sn-intro01} also explains that it can
be comprised of different sensor devices playing different roles in what they
observe in the environment, characterizing the heterogeneity of the sensor
network by the hardware capabilities and what types of data each of them
produce.

Finally, \cite{sn-intro01} mentions that sensor device nodes can communicate
with each other using different media such as radio, diffuse light or
inductive sound, influencing the type of protocols they use to exchange data. They
can have knowledge of their physical location if they are integrated with a
Global Positioning System (GPS) device, which gives the latitude or
longitude coordinate values. Therefore, the cost of these devices may range
from hundreds to thousands of dollars, depending on the design capabilities and functionality of a given device.

@@

The last section discussed the different properties of the sensor devices when
placed in a sensor network. Depending on how they communicate with each other,
\cite{sn-intro02} states that sensor devices can be organized in a
structured, or ad hoc way. For this reason, the infrastructure of sensor
networks differ in topology such as single-hop, star, tree or  graph. Each of
these topologies may influence how the produced data from sensor nodes is
collected and routed to their related data sink. While there are only two
levels of node connectivity in sensor networks deployed as single-hop and
star topologies, sensor networks whose topology is either tree or graph use
multi-hop communication among each of their nodes for data communication from
the devices to the data sink.

The property of coverage on a sensor network is defined by
\cite{sn-intro01} as the physical space ``visible" to the sensors devices, which
is also directly related to the purpose of the network, its size in terms of
the number of nodes deployed and the sensor network lifetime. Therefore,
sensors are deployed in sparse or dense manners, according to the
specification of what needs to be covered by the sensor devices. Furthermore,
considering that some types of sensor devices are very likely to experience regular
hardware failures, \cite{sn-intro02} described sensor network examples with 
redundant nodes that provide better data coverage.
Finally, the coverage of sensor devices also have influence on the connectivity
property of the sensor nodes. In order to communicate with redundant nodes, a
physical connection must exist at all times. On the other hand, lower
level of coverage is related to sporadic communication among the sensor nodes.

The sensor network size is related to the degree of coverage mentioned earlier,
and it is measured by the number of devices deployed. \cite{sn-intro01}
reports examples of sensor networks composed by different ranges of deployed
sensor devices such as dozens, hundreds or even thousands of nodes. However,
the lifetime of a sensor network determines how long these devices are going
to be used, as well as how long the sensor network may exist. For example, an
environmental sensor network \cite{sn-ex01} with the purpose of collecting data
for historic purposes may have their sensor devices continuously performing
their data collection tasks. On the other hand, the lifetime of a
sensor network designed to achieve a specific goal will not exist after the sensor
devices finish their tasks. Finally, \cite{sn-intro02} argues that the
performance related to the orchestration of sensor devices in sensor networks
is related to the constraints defined for quality of service (QoS).  This includes
real-time data delivery within a threshold or for robustness such as
maintaining live data even though a communication link is not available.

All things considered, sensor nodes play a vital role during the execution
of a sensor network, whose infrastructure is characterized by how and where the
sensor nodes are deployed and communicate with each other. The next
section covers the approaches dealing with the problem of this type of system
network.

@@

As outlined in the last few sections, sensor networks are comprised of
different types of sensor nodes. An important sensor node type is the
data sink because it is where the raw data produced by sensor
nodes resides. Data collection and access in sensor networks is directly
related to the characteristics of the sensor network, its routing mechanisms
and where the data sink is placed within the network.

The type of query mechanism used to directly or indirectly interrogate sensor
devices in a sensor network is directly related to the aforementioned locations
of the collected data. The in-network query processing is used when the
location of the collected data is a local storage site \cite{sn-storage01,
sn-storage03} within the network; a strategy typically used to maximize the
sensor's resources, while minimizing energy consumption \cite{sn-storage04}. On
the contrary, the a typical centralized query processing is used to search for
values over the collected data when the location is a centralized data
node \cite{sn-storage02}.

When it comes to the data representation, the relational data model
\cite{relational-model} is one of the most traditional approaches used to
describe the collected data of sensor networks, as it is reported in database
systems such as the TinyDB \cite{sn-db-tinydb}. For instance, \cite{sn-db-newop}
proposes a modified version of the SQL language of TinyDB proposed to solve
SQL Join problems, so that the query processing takes into account the location
of the distributed data. Conversely, \cite{sn-provenance} proposes the use of
Data Provenance as the primary guidelines used to describe collected  raw data from
sensor devices.

@@

As presented in the previous section, the raw data is the one in-transit from
the sensor to the storage device, where it is labeled to be
identifiable. It is also referred to as the collected data. 
Initially, the raw data serves as the main outcome of the sensor network.
\cite{sn-provenance} differentiates the use of \textbf{real-time data stream}, when they are generated from a sensor
device whenever the data is requested. In this way, the data is still in its raw
format specified by the sensor manufacturer. However, it is not retained. 
In contrast, the authors describe the data as \textbf{archivable}, when they a
have historical significance and will be processed to be reused by other systems at
a latter time. While the real-time data streams are used and potentially
discarded, the archivable data is stored into secondary memory such as a
hard-disk.

After producing the ``sensed'' values for the observed parameters from the
environment, the sensor node is responsible for sending the raw data to a specific
data node defined by the network infrastructure, as delineated in section
\ref{sec:sn-infrastructure}. Thus, \cite{sn-storage03}
describes that the destination for the collected sensor data will be one of two
different types of storage devices: the \textbf{external storage} and the
\textbf{local storage}. In the former type, the sensor sends the
produced data to another sensor node that contains an external storage device.
In the latter the data is stored only within the memory of the device before its 
reading as a real-time data stream. On the whole, the sensor network
infrastructure dictates where the movement performed by the sensor data,
for example, \cite{sn-storage02} sees a wireless sensor networks as a many-to-one data
collection pattern, when its organization is based on single-hop or star. On
the other hand, data may also flow from one intermediate sensor node to another
until it reaches its final data sink when the data organization is structured
as a graph, tree \cite{sn-storage01, sn-storage03}.

Lastly, the use of different external storage devices can be used
to partition the collected data. This practice is called \textbf{Data-Centric
Storage}, and aims at clustering the collected data from different sensor
devices based on a given a property of the data such as location or time of
creation.

@@

The query processing in sensor networks is a direct or indirect method of
accessing the collected data stored at data nodes, whose location is determined
by different facts described in the previous section. \cite{sn-storage03}
describes an example of an \textbf{in-network query processing} when sensor
nodes, acting as data nodes, provides real-time data stream as a response
to queries requesting the device's collected data. Likewise, if the collected
data is in use for a specific period of time, the data is cached in an external
storage device in a close-by data node. In this way, when a query is issued to
the sensor network with these characteristics, the query processor considers 
all data nodes before it returns its final response to the query. In comparison, a sensor
network whose topology is in the form of a star, the data sink is usually
located in the center of the structure, and thus, receives all
the collected data from other sensor devices within the network. For this reason, it 
can be seen as a \textbf{centralized query processing}. Factoring in that a centralized data
sink receives all the collected data from the peer nodes, this query processing
strategy is responsible for responding to indirect queries about the
produced data from the network.

Based on the literature review, there are advantages and disadvantages related to
each of the query processing approaches described with regards to the
infrastructure of the sensor network. \cite{sn-storage03} argues that the 
primary purpose of the in-network query processing is to minimize the amount
of data transmitted from sensor devices to a data sink, as sensor devices are
generally limited by energy use. Considering that sensor devices are also used
as data nodes, \cite{sn-storage04} states that this approach can potentially
flood all sensor nodes in the network, thereby decreasing its performance.
For this reason, \cite{sn-storage01} reports the use of the data-centric
storage by organizing the collected data based on its particular property
values. In this way, the query processor hits specific data nodes that contain
the types of data in the range of the requested data concluding that this approach 
decreases the network congestion and maximizes the
capacity on the sensor devices and the performance of the sensor network.

Unlike the in-network, the centralized query processing is related to a regular
database system, located in one of the network nodes. For this reason, this
indirect type of query processing is used when the data purpose of the data is for 
data storage. As a consequence, \cite{sn-storage02} shows that the result of
a many-to-one communication between the centralized data sink
and the other data nodes, this query process does not affect any of the sensor
nodes in the network and, thus, maximizes the sensor device utilization. 
Moreover, this strategy better exposes the collected data to the primary users
raw data, which follows the process of data processing, compression,
aggregation, and so on \cite{sn-db-modeling02}.

Although the approach of centralized query processing maximizes the sensor
nodes capacity and minimize energy consumption, \cite{sn-storage02} explains
that this trade off imposes the creation of the unavoidable development of a
point of traffic concentration or the phenomenon of network congestion. For
this reason, \cite{sn-storage04} describes that different techniques used in 
distributed systems such as the use of master-slave data replication or 
database partitioning \cite{db-partitioning-relational} can be used to reduce
or mitigate the effects of this funneling.

@@

Different approaches are used to represent the collected data from sensors.
Some research groups lacking database experience arbitrarily assign data 
descriptors, while others use database technology in a more pragmatic approach.
\cite{sn-programming-language} describes several applications using
various types of languages for accessing the data designed in different data
models. \cite{sn-provenance} reports on the use of a \textbf{Tabular Data Model}
\cite{tabular-model} using comma-separated values (CSV) to organize the
measurements collected from sensor devices. In addition, \cite{sn-db-tinydb} 
describes the use of the \textbf{Relational Data Model} \cite{relational-model} 
as used in the majority of the projects that involves data persistence due to its
popularity in areas such as Software Engineering and Distributed Systems. Finally,
\cite{sn-xml-usage01, sn-xml-usage02} discusses examples of sensor network 
applications that make use of XML \cite{xml} to represent the collected data on a 
\textbf{Structured Data Model}.

@@

The most basic computer application to process data is through the use of
spreadsheets, where data is organized into cells \cite{tabular-model}. When
exported to the so-called comma-separated values (CSV) file format, the
resulting file is comprised of lines of ASCII characters, separated by a
delimiter. The first line represents a header, with the identification of
columns of specific data, where the following lines contains the values
related to each top column. \cite{sn-provenance} reports that the tabular
data model is commonly used to provide data persistence for sensor networks.
The SF-BEAMS environmental sensor networks \cite{sfbeams2006} uses this
approach to distribute the collected data over the Internet using the OPenDAP
\cite{opendap} data format.

This data model can be used in different contexts, as \cite{sn-provenance}
describes that the versions of the same document can be easily managed and
shared among research colleagues using a version-control system such as
Subversion \cite{subversion}.

@@

\cite{sn-db-newop} shows the use of the Relational Data Model
\cite{relational-model} as it is used in sensor networks to describe the
collected data from sensor devices. The project described
by the authors is based on the TinyDB database system \cite{sn-db-tinydb}, a
database system specifically built for data nodes of sensor networks. It uses
the same properties and constraints of regular database systems, where the
entities of the sensor network are required to be modeled and represented by
tables.

@@

XML is an increasing popular data format used widely by the applications developed
for the Internet. \cite{sn-xml-usage01} reports an efficient way to represent
the collected data from sensors using XML, while the authors of
\cite{sn-xml-usage02} describe a sensor network using XML for data model
representation. In general, the collected data is structured into classes of
relating elements and validated by an XML Schema \cite{xml-schema}.

@@

Other examples of the use of the structured data model using XML is found in
the use of middlewares, as the authors of \cite{sn-xml-query-engines} reports.
Such an approach is used for data exchange by the implementing of a middleware
capable of query and extract the collected data from sensor devices
\cite{sn-xml-middleware}. A practical example of the use of XML as a middleware
to communicate between nodes is the communication mechanism used by NetBEAMS
\cite{netbeams2009}. See section \ref{sec:dsp-serialization} for details.

\section{Description of the Collected Data}
\label{sec:sn-data-description}

As described in the previous section, the collected data from sensor networks
can be represented by different data models, requiring the knowledge of the
properties of the data. Not to mention that the low-level or raw data produced
by a given sensor device might not include vital details regarding the
origin and identity of the collected data. For this reason, the collected data
is reported to be described in different ways such as Data Provenance or
Annotations.

\cite{sn-provenance} describes Data Provenance as a technique used to describe
those essential properties of the collected data from sensors. One typical
example of the application of Data Provenance is the so-called black-box of an
aircraft, which carries the collection of spatial-temporal data structures
collected from a group of sensors placed in different locations of an
aircraft, as shown by \cite{sn-exemple-blackbox}. Finally, the author describes 
the collected data as used in the process of data recovery and analysis in order to
identify the probable reason of an aircraft accident, by analyzing the time, 
location of the aircraft, and other independent variables from the sensors.

In general, sensor devices produce and preserve data onto their local
storage or transmit them to data nodes as described earlier in this chapter. It
is vital to understand that the majority of these devices outputs the raw data
in streams of values with any type of delimitation between each other. For
this reason, the processes of parsing and labeling each of the values located
in the data stream regularly requires the use of the device's specification
documentation. Similarly, in order to have access to each of the properties of
collected data by a given property related to time or space,
\cite{sn-provenance} also mentions that Metadata\footnote{Data about data}
should be included together with the collected data.

An example of a Data Provenance application is the use of the aggregation of data
over time to estimate the changing effects of the weather in a given region.
For this reason, the collected data from sensor devices need meaningful
annotations to better describe the identity of the collected data, as 
\cite{sn-provenance} reports. Pertinent questions regarding the
data can be asked such as \textbf{what was collected?}, \textbf{When was the
data collected?} or \textbf{Where was the data collected?}. Similarly, while
Data Provenance provides basic mechanisms to describe sensor networks data in
general, others see the importance of the time and position of sensor networks.
\cite{sn-time-series} describes strategies used to cluster distributed sensor
data as time series of data, as well as to decrease the size of data streams
being transmitted in real-time sensor networks when these collected data do
not need to be transferred \cite{sn-data-reduction}.

A different way to describe sensor devices is through the use of
Annotations. \cite{sn-annotation} shows the use of tags to describe data
collected from a physical area monitoring. The sensor network was
designed to monitor moving objects, capturing their location while in
movement and their visual appearance using a video camera. In this way,
different instances can be described by adding tags to particular video frames, and
therefore, providing an efficient way to search for a specific frame based on
its visual observations. Similarly, the above referenced project also uses data
provenance to describe its collected data, since each of the collected data 
contains the related GPS coordinates of the observed object.

Finally, data provenance provides practical use of the collected data from
different sensor devices and different sensor networks in the process called
Information Fusison \cite{sn-info-fusion}. \cite{sn-geo-metadata} is an example of
the aggregation of different data from different devices located in different 
geographic sensor networks which is capable of processing spatial-temporal 
queries of real-time or historical data. The following sections describe each 
of the metadata properties as defined by \cite{sn-provenance}, giving different
examples of each of them.

\subsection{Data as an object}

Different sensor devices produce different sets of attributes based on their
capabilities and design. In general, these data sets do not have distinct
names, which leads to difficulties when indices over the properties of the
collected data are required to make them searchable. Specially for the purpose
of data archival, the collected data needs to be identified, giving a meaning
of independence on its own, the same idea behind the unique identification of
entities on the relational data model \cite{relational-model}. Equally,
\cite{sn-provenance} describes that traditional sensor networks use the
collected data for analysis long after the attributes have been collected,
usually before it is categorized and indexed. Annotations and tags are also
names given by \cite{sn-provenance} as means to describe collected sensor data.

Briefly, the collected data is described by the name of the observation
property. For instance, \textbf{temperature} = 73.45 expresses the value of the
observation of the temperature. In other words, the string ``temperature'' is
the property describing the raw collected data 73.45. On the other hand,
the example \textbf{temperature-scale} = ``Fahrenheit": describes the metadata
regarding the temperature.

\subsection{Time of Data Creation}
